---
title: "Module 1: Mastering R for Spatial Data"
subtitle: "A Structured Coursebook"
---

```{r setup, include=FALSE}
# This code chunk sets up the global options for all subsequent code chunks.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

The goal of this module is to build a rock-solid foundation in handling, manipulating, and visualizing spatial data within the R environment. We will focus on the "tidyverse" approach to spatial analysis, which treats spatial data as a special kind of data frame, making many operations intuitive and powerful.

By the end of this notebook, you will be able to:

1.  Set up your R environment by loading the essential packages for geospatial analysis.
2.  Understand, load, inspect, and manipulate **vector data** using the `sf` package.
3.  Understand, load, and perform fundamental operations on **raster data** using the `terra` package.
4.  Create beautiful and informative static maps using the `tmap` and `ggplot2` packages.
5.  Generate interactive maps for data exploration using `tmap`'s "view" mode.

------------------------------------------------------------------------

## 1.1 Setup: Installing and Loading Core Packages

For our work, we will rely on a core set of packages that form the backbone of modern spatial analysis in R.

If you have not installed these packages before, you must do so first. You can run the `install.packages()` function in your R console for each one. You only need to do this once.

```{r install_code, eval=FALSE, echo=TRUE}
# Run these lines in your R console if you haven't installed the packages yet
# install.packages("sf")
# install.packages("terra")
# install.packages("tmap")
# install.packages("ggplot2") # The premier package for data visualization
# install.packages("dplyr")
# install.packages("spData")
# install.packages("geodata")
# install.packages("spDataLarge", repos = "https://nowosad.github.io/drat/", type = "source")
```

Once installed, we need to load the packages into our current R session using the `library()` function. We do this at the start of every script.

```{r load_packages}
# --- Core Packages ---
library(sf)         # Handles vector data (Simple Features). The modern standard.
library(terra)      # Handles raster data. The modern, high-performance successor to the 'raster' package.
library(tmap)       # Used for creating beautiful thematic maps (like ggplot2 for maps).
library(ggplot2)    # A powerful and versatile grammar of graphics for visualization.
library(dplyr)      # A grammar of data manipulation, works seamlessly with 'sf'.

# --- Data Packages ---
library(spData)     # Contains example spatial datasets for practice.
library(geodata)    # For downloading common global spatial datasets like country boundaries or climate data.
```

## 1.2 Vector Data with `sf`

### What is Vector Data?

Vector data represents geographical features using discrete geometric objects: **Points**, **Lines**, and **Polygons**.

### The `sf` Object

The `sf` package provides a framework for working with vector data in R. An `sf` object is fundamentally an R **data frame** with a special list-column that stores the geometry. This structure is powerful because it allows you to use `dplyr` verbs directly on your spatial data.

### Worked Example: Exploring World Countries

Let's load the `world` dataset from the `spData` package and see what an `sf` object looks like.

```{r sf_worked_example}
# Load the 'world' dataset
data(world)

# 1. Inspect the object's class
class(world)

# 2. Print the object
print(world)

# 3. Manipulate data with dplyr
world_africa <- world %>%
  filter(continent == "Africa") %>%
  select(name_long, pop, gdpPercap, lifeExp, geom)

# Let's look at our new, smaller sf object
print(world_africa)

# 4. Check the Coordinate Reference System (CRS)
st_crs(world_africa)

# 5. Basic Plotting
plot(st_geometry(world_africa), main = "Map of Africa (Geometries Only)")
```

------------------------------------------------------------------------

### **Exercise 1: Vector Data Manipulation**

Now it's your turn. Use the `us_states` dataset, which is also included in the `spData` package.

::: panel-tabset
#### Task 1

Load the `us_states` data and inspect its Coordinate Reference System (CRS).

#### Try it yourself!

```{r ex1_task1_prompt}
# Your code here
```

#### Solution

```{r ex1_task1_solution}
data(us_states)
st_crs(us_states)
```
:::

::: panel-tabset
#### Task 2

Use `dplyr::filter()` to create a new `sf` object containing only the states in the "West" region.

#### Your turn!

```{r ex1_task2_prompt}
# Your code here
```

#### Solution

```{r ex1_task2_solution}
western_states <- us_states %>%
  filter(REGION == "West")

# Optional: Print the names to check your work
print(western_states$NAME)
```
:::

::: panel-tabset
#### Task 3

Calculate the area of each western state using `st_area()`. The result will be in square meters. Create a new column in your `western_states` object called `area_sqkm` that contains the area in square kilometers (Hint: 1 $km^2$ = 1,000,000 $m^2$ ).

#### Try it yourself!

```{r ex1_task3_prompt}
# Your code here
```

#### Solution

```{r ex1_task3_solution}
# First, create the 'western_states' object from the previous task
western_states <- us_states %>% 
  filter(REGION == "West")

# Now, add the new area column
western_states_with_area <- western_states %>%
  mutate(area_sqkm = st_area(.) / 1000000)

# View the NAME and the new area column
print(western_states_with_area[, c("NAME", "area_sqkm")])
```
:::

::: panel-tabset
#### Task 4

Create a simple plot of your `western_states` object using `plot()`. Customize it by changing the color of the polygons (`col`) and their borders (`border`).

#### Try it yourself!

```{r ex1_task4_prompt}
# Your code here
```

#### Solution

```{r ex1_task4_solution}
# First, create the 'western_states' object from task 2
western_states <- us_states %>% 
  filter(REGION == "West")

plot(st_geometry(western_states), 
     main = "Western US States", 
     col = "khaki", 
     border = "darkgreen")
```
:::

------------------------------------------------------------------------

## 1.3 Raster Data with `terra`

### What is Raster Data?

While vector data uses discrete shapes, raster data represents the world as a continuous grid of cells, or **pixels**. Each pixel in the grid has a value representing some measured phenomenon. Think of it like a digital photograph.

*Epidemiological examples: Satellite imagery, elevation models, or surfaces showing environmental variables like land surface temperature, precipitation, or vegetation density (NDVI).*

### The `terra` Package

The `terra` package is the modern, high-performance engine for raster data in R. It is designed to be easier to use and much faster than its predecessor, the `raster` package, especially with the very large files common in remote sensing.

### Worked Example: Exploring Elevation in Uganda

Let's download elevation data for Uganda and perform two fundamental raster operations: **cropping** and **masking**.

```{r terra_worked_example}
# We will use the sf object for Uganda we can get from the 'world' dataset
uganda_sf <- world %>% filter(iso_a2 == "UG")

# 1. Download Data
# The geodata package can download various global datasets.
# We will get elevation data at 30-arcsecond resolution.
temp_path <- tempdir()
elevation_uganda <- geodata::elevation_30s(country = "UGA", path = temp_path)

# 2. Inspect the SpatRaster Object
print(elevation_uganda)

# 3. Basic Plotting
plot(elevation_uganda, main = "Elevation of Uganda (Rectangular Extent)")
plot(st_geometry(uganda_sf), add = TRUE, border = "red", lwd = 2)

# 4. Crop and Mask
elevation_cropped <- crop(elevation_uganda, uganda_sf)
elevation_masked <- mask(elevation_cropped, uganda_sf)

# 5. Plot the Final Result
plot(elevation_masked, main = "Masked Elevation of Uganda (m)")
plot(st_geometry(uganda_sf), add = TRUE, border = "black", lwd = 1)
```

------------------------------------------------------------------------

### **Exercise 2: Raster Data Manipulation**

Let's practice these skills by creating a map of the mean annual temperature for Brazil.

::: panel-tabset
#### Task 1

Download the global average maximum temperature data from WorldClim at 10-minute resolution. This will be a `SpatRaster` with 12 layers (one for each month). Calculate the mean across all 12 layers to get a single-layer raster of the annual average. *Remember: The data is in °C \* 10, so you'll need to divide the final result by 10.*

#### Try it yourself!

```{r ex2_task1_prompt}
# Your code here
```

#### Solution

```{r ex2_task1_solution}
temp_path <- tempdir() # Define temp_path if not already defined
global_tmax <- geodata::worldclim_global(var = "tmax", res = 10, path = temp_path)
mean_annual_tmax <- mean(global_tmax)
mean_annual_tmax_c <- mean_annual_tmax / 10

# Optional: print the final raster object to check it
print(mean_annual_tmax_c)
```
:::

::: panel-tabset
#### Task 2

Create an `sf` object for Brazil from the `world` dataset.

#### Try it yourself!

```{r ex2_task2_prompt}
# Your code here
```

#### Solution

```{r ex2_task2_solution}
brazil_sf <- world %>%
  filter(name_long == "Brazil")
  
print(brazil_sf)
```
:::

::: panel-tabset
#### Task 3

Crop and then mask the global temperature raster (`mean_annual_tmax_c`) to the exact shape of Brazil.

#### Try it yourself!

```{r ex2_task3_prompt}
# Your code here
```

#### Solution

```{r ex2_task3_solution}
# Re-create objects from previous tasks to make this chunk self-contained
temp_path <- tempdir()
global_tmax <- geodata::worldclim_global(var = "tmax", res = 10, path = temp_path)
mean_annual_tmax_c <- mean(global_tmax) / 10
brazil_sf <- world %>% filter(name_long == "Brazil")

# Now perform the crop and mask
tmax_brazil_cropped <- crop(mean_annual_tmax_c, brazil_sf)
tmax_brazil_masked <- mask(tmax_brazil_cropped, brazil_sf)

print(tmax_brazil_masked)
```
:::

::: panel-tabset
#### Task 4

Plot your final masked temperature raster for Brazil. Add the country's border to the plot for context.

#### Try it yourself!

```{r ex2_task4_prompt}
# Your code here
```

#### Solution

```{r ex2_task4_solution}
# Re-create the necessary objects from previous tasks
temp_path <- tempdir()
global_tmax <- geodata::worldclim_global(var = "tmax", res = 10, path = temp_path)
mean_annual_tmax_c <- mean(global_tmax) / 10
brazil_sf <- world %>% filter(name_long == "Brazil")
tmax_brazil_masked <- mask(crop(mean_annual_tmax_c, brazil_sf), brazil_sf)

# Now create the plot
plot(tmax_brazil_masked, main = "Mean Annual Max Temperature for Brazil (°C)")
plot(st_geometry(brazil_sf), add = TRUE, border = "black")
```
:::

------------------------------------------------------------------------

## 1.4 Creating Maps with `tmap`

While `plot()` is great for quick checks, the `tmap` package provides a powerful and flexible system for creating publication-quality thematic maps. It uses a "grammar of graphics" approach, similar to `ggplot2`, where you build a map layer by layer.

The main components are:
* `tm_shape()`: Specifies the `sf` or `SpatRaster` object to be mapped.
* `tm_polygons()`, `tm_lines()`, `tm_dots()`, `tm_raster()`: Specify *how* to draw the shape.
* `tm_layout()`, `tm_compass()`, `tm_scale_bar()`: Functions to customize the map's appearance and add map furniture.

A key feature of `tmap` is its dual-mode functionality. You can create:
1.  **Static Maps (`tmap_mode("plot")`)**: High-quality images suitable for publications, reports, and presentations.
2.  **Interactive Maps (`tmap_mode("view")`)**: Dynamic maps that you can pan, zoom, and query in a web browser or the RStudio Viewer. This is incredibly useful for data exploration.

### Worked Example: Static Maps

Let's create a high-quality map showing life expectancy in Africa, highlighting Uganda, and another map showing Uganda's elevation.

```{r tmap_worked_example, fig.show='hold'}
# Set tmap mode to "plot" for static maps.
tmap_mode("plot")

# Map 1: A Choropleth Map of Life Expectancy
# A choropleth map is one where polygons are colored according to a variable's value.
map1 <- tm_shape(world_africa) +
  tm_polygons(
    col = "lifeExp",                      # The column to map
    palette = "plasma",                   # Color palette
    style = "quantile",                   # How to break data into color bins
    title = "Life Expectancy (Years)"     # Legend title
    ) +
  tm_shape(uganda_sf) +                   # Add a new shape layer for Uganda
  tm_borders(col = "black", lwd = 2.5) +  # Draw its borders
  tm_layout(
    main.title = "Life Expectancy in African Nations",
    main.title.position = "center",
    legend.position = c("left", "bottom"),
    frame = FALSE, 
    inner.margins = c(0.05, 0.1, 0.05, 0.05)
    ) +
  tm_compass(type = "8star", position = c("right", "top"), size = 3) +
  tm_scale_bar(position = c("right", "bottom"), breaks = c(0, 1000, 2000))

# Map 2: Combining Raster and Vector Data
# Let's map the masked Uganda elevation raster we created earlier.
map2 <- tm_shape(elevation_masked) +
  tm_raster(
    title = "Elevation (m)"
    ) +
  tm_shape(uganda_sf) +
  tm_borders(lwd = 1.5, col = "black") +
  tm_layout(main.title = "Elevation of Uganda")  +
  tm_compass(type = "arrow", position = c("left", "top")) +
  tm_scale_bar()

# Print the maps
map1
map2
```

### The Power of Interactive Mode

The real magic of `tmap` for data exploration comes from its interactive mode. By simply running `tmap_mode("view")` once, all subsequent `tmap` plots become interactive Leaflet maps. You can click on features to see their data, zoom in on areas of interest, and switch between different basemaps. This is invaluable for sanity-checking your data and discovering patterns.

Let's take our first map and view it interactively.

```{r tmap_interactive_example}
# Switch to interactive view mode
tmap_mode("view")

# Re-run the same code for our first map
# No other changes are needed!
tm_shape(world_africa) +
  tm_polygons(
    fill = "lifeExp",
    # When hovering, show the country name and its life expectancy
    popup.vars = c("Country" = "name_long", "Life Expectancy" = "lifeExp")
    ) +
  tm_shape(uganda_sf) +
  tm_borders(col = "black", lwd = 2.5) +
  tm_layout(main.title = "Interactive Map of Life Expectancy in Africa")

# --- IMPORTANT ---
# Switch back to plot mode for the rest of the document
tmap_mode("plot")
```

------------------------------------------------------------------------

### **Exercise 3: Thematic Mapping with `tmap`**

Now, create your own high-quality map of the United States.

::: panel-tabset
#### Task 1

Using the `us_states` dataset, create a choropleth map showing the total population in 2015 (`total_pop_15`). Add state borders with `tm_borders()`.

#### Try it yourself!

```{r ex3_task1_prompt}
# Your code here
```

#### Solution

```{r ex3_task1_solution}
# Ensure us_states data is loaded
data(us_states)

tm_shape(us_states) +
  tm_polygons(col = "total_pop_15", title = "Population (2015)") +
  tm_borders(col = "gray70")
```
:::

::: panel-tabset
#### Task 2

Improve the map from Task 1.
* Add a main title.
* Change the color palette (try `"YlGnBu"` or `"Reds"`).
* Add a compass and a scale bar.
* Make the state borders white and thin (`lwd = 0.5`).

#### Try it yourself!

```{r ex3_task2_prompt}
# Your code here
```

#### Solution

```{r ex3_task2_solution}
# Ensure us_states data is loaded
data(us_states)

tm_shape(us_states) +
  tm_polygons(
    col = "total_pop_15", 
    title = "Total Population (2015)",
    palette = "YlGnBu",
    style = "jenks" # Jenks style is good for skewed data
    ) +
  tm_borders(col = "white", lwd = 0.5) +
  tm_layout(
    main.title = "Population of the United States, 2015",
    frame = FALSE,
    legend.outside = TRUE
    ) +
  tm_compass(type = "arrow", position = c("right", "top")) +
  tm_scale_bar(position = c("left", "bottom"))
```
:::

------------------------------------------------------------------------

## 1.5 Creating Maps with `ggplot2`

For those already familiar with the `ggplot2` package, its "grammar of graphics" can be extended to create highly customized, professional maps. The `sf` package integrates directly with `ggplot2` through a special geometric layer: `geom_sf()`.

The core idea is the same as any other ggplot: you initialize a plot with `ggplot()`, specify your data, and then add layers. For spatial data, `geom_sf()` is the primary layer.

### Worked Example: Mapping US Population with `ggplot2`

Let's use `ggplot2` and `geom_sf` to create a map of population density for the `us_states` dataset.

```{r ggplot_worked_example}
# Load the data if not already in the environment
data(us_states)

# We can add a new column for population density right in our pipe
us_states_plot <- us_states %>%
  mutate(pop_density = as.numeric(total_pop_15 / AREA))

# Create the map with ggplot2
ggplot(data = us_states_plot) +
  # The main geometry layer for sf objects
  geom_sf(aes(fill = pop_density)) + 
  
  # Customize the color scale
  scale_fill_viridis_c(
    trans = "log10", # Use a log scale for better visualization of skewed data
    name = "Population Density\n(people / sq. mile, log scale)"
    ) +
  
  # Add titles and a clean theme
  labs(
    title = "US Population Density by State, 2015",
    subtitle = "Data from the spData package",
    caption = "Map created with ggplot2"
  ) +
  theme_void() + # A minimal theme, good for maps
  theme(legend.position = "bottom")

```

### **Exercise 4: Thematic Mapping with `ggplot2`**

Let's use `ggplot2` to visualize the economic data in the `world_africa` dataset we created earlier.

::: panel-tabset
#### Task 1

Using the `world_africa` object, create a simple map of the African continent using `ggplot()` and `geom_sf()`.

#### Try it yourself!

```{r ex4_task1_prompt}
# Your code here
```

#### Solution

```{r ex4_task1_solution}
# Assuming 'world_africa' was created in section 1.2
ggplot(data = world_africa) +
  geom_sf() +
  ggtitle("Map of Africa")
```
:::

::: panel-tabset
#### Task 2

Map the `gdpPercap` (GDP per capita) variable to the `fill` aesthetic of the polygons.

#### Try it yourself!

```{r ex4_task2_prompt}
# Your code here
```

#### Solution

```{r ex4_task2_solution}
ggplot(data = world_africa) +
  geom_sf(aes(fill = gdpPercap))
```
:::

::: panel-tabset
#### Task 3

Improve your map.
* Use `scale_fill_viridis_c()` for a better color scale and add a title to the legend.
* Add a main title and a subtitle to the map using `labs()`.
* Apply `theme_bw()` for a cleaner look.
* Add a black border layer for Uganda (`uganda_sf`) on top of the other countries. (Hint: Use a second `geom_sf()` call, but set `fill = NA` so it doesn't cover the colors).

#### Try it yourself!

```{r ex4_task3_prompt}
# Your code here
```

#### Solution

```{r ex4_task3_solution}
# Assuming 'world_africa' and 'uganda_sf' are loaded
ggplot() +
  # Layer 1: African countries colored by GDP
  geom_sf(data = world_africa, aes(fill = gdpPercap)) +
  
  # Layer 2: Uganda outline
  geom_sf(data = uganda_sf, fill = NA, color = "red", linewidth = 0.8) +
  
  # Customize scales and labels
  scale_fill_viridis_c(name = "GDP per Capita (USD)") +
  labs(
    title = "GDP Per Capita in Africa",
    subtitle = "Highlighting Uganda",
    caption = "Source: spData package"
  ) +
  theme_bw()
```
:::

------------------------------------------------------------------------

# Module 1 Conclusion

Congratulations on completing the first module!

You have acquired the fundamental skills for working with spatial data in R. You have learned how to use `sf` and `dplyr` to manage and query vector data, how to use `terra` to perform essential operations on raster data, and how to turn that data into informative and visually appealing maps.

Specifically, you now have two powerful mapping tools at your disposal:
* **`tmap`**, which provides a straightforward grammar for creating both static, publication-quality maps and incredibly useful interactive maps for data exploration.
* **`ggplot2`**, which extends its famous grammar of graphics to spatial data with `geom_sf`, allowing for deep customization and integration into a tidyverse workflow.